---
title: "Grammar of graphic-ggplot2"
author: "Minh Trung DANG"
date: "23/05/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r cars}
qplot(displ, hwy, data=mpg, facets = . ~ year) + geom_smooth()
```


Scaling actually occurs in three parts: transforming, training and mapping.

 Scale transformation occurs before statistical transformation so that statis-
tics are computed on the scale-transformed data. This ensures that a plot
of log(x) vs. log(y) on linear scales looks the same as x vs. y on log scales.
There are many different transformations that can be used, including taking
square roots, logarithms and reciprocals.

 Scale transformation occurs before statistical transformation so that statis-
tics are computed on the scale-transformed data. This ensures that a plot
of log(x) vs. log(y) on linear scales looks the same as x vs. y on log scales.
There are many dierent transformations that can be used, including taking
square roots, logarithms and reciprocals.

Finally the scales map the data values into aesthetic values. This is a
local operation: the variables in each dataset are mapped to their aesthetic
values producing a new dataset that can then be rendered by the geoms.

the components that make up a plot (a layer): 
  data 
  aesthetic mappings, 
  geometric objects (geoms), 
  statistical transformations (stats), 
  scales
  faceting


A plot may have multiple layers, as in the example where we overlaid a smoothed line on a scatterplot.

All together, the layered grammar defines a plot as the combination of:

** A default dataset and set of mappings from variables to aesthetics.
** One or more layers, each composed of a geometric object, a statistical transformation, and a position adjustment, and optionally, a dataset and
aesthetic mappings.
** One scale for each aesthetic mapping.
** A coordinate system.
** The faceting specification.

```{r}
p <- qplot(displ, hwy, data = mpg, colour = factor(cyl))
p
```


```{r}
summary(p)
```

# Save plot object to disk
save(p, file = "plot.rdata")
# Load from disk
load("plot.rdata")

# Save png to disk
```{r}
ggsave("plot.png", width = 5, height = 5)
```

Build a plot layer by layer

Each layer can come from a dierent dataset and have a dierent aesthetic mapping, allowing us to create plots that could not be generated
using qplot(), which permits only a single dataset and a single set of aesthetic mappings.


1. The data  which must be an R data frame, and can be changed after the plot is created.
2. A set of aesthetic mappings, which describe how variables in the data are mapped to aesthetic properties of the layer. 
3. This section includes a description of how layer settings override the plot defaults, the difference between setting and mapping, and the important group aesthetic.
4. The geom, which describes the geometric used to draw the layer. The geom defines the set of available aesthetic properties.
5. The stat, which takes the raw data and transforms it in some useful way. 
The stat returns a data frame with new variables that can also be mapped to aesthetics with a special syntax.
6. The position adjustment, which adjusts elements to avoid overplotting.


Creating a plot
```{r}
p <- ggplot(diamonds, aes(carat, price, color =  cut))
p
```

This plot object cannot be displayed until we add a layer: there is nothing to see!


4.3 Layers

```{r}
p2 <- ggplot(diamonds, aes(x = carat))
p2 <- p2 + geom_histogram(binwidth = 2, fill = "steelblue")
p2
```


All the shortcut functions have the same basic form, beginning with geom_ or stat_:
  geom_XXX(mapping, data, ..., geom, position)
  stat_XXX(mapping, data, ..., stat, position)
  
  
***mapping (optional): A set of aesthetic mappings, specified using the aes() function and combined with the plot defaults
***data (optional): A dataset which overrides the default plot dataset. 
It is most commonly omitted, in which case the layer will use the default plot data.
***...: Parameters for the geom or stat, such as bin width in the histogram or bandwidth for a loess smoother. 
You can also use aesthetic properties as parameters. When you do this you set the property to a fixed value, not
map it to a variable in the dataset. The example above showed setting
the fill colour of the histogram to steelblue".

***geom or stat (optional): You can override the default stat for a geom, or the default geom for a stat. 
This is a text string containing the name of the geom to use. Using the default will give you a standard plot; 
overriding the default allows you to achieve something more exotic,

***position (optional): Choose a method for adjusting overlapping objects,


NOTE : the order of data and mapping arguments is switched between ggplot() and the layer functions.

In ggplot : ggplot(data = diamonds, aes(x = carat))
In lyaer function: geom_XXX(mapping, data, ..., geom, position)

Because you almost always specify data for the plot, and almost always specify aesthetics|but not data|for the
layers.

```{r}
p5A <- ggplot(msleep, aes(sleep_rem /sleep_total, awake)) +
geom_point()


p5B <- qplot(sleep_rem / sleep_total, awake, data = msleep)

ggarrange(p5A, p5B, ncol = 2, nrow = 1)

```


```{r}
p6A <- qplot(sleep_rem / sleep_total, awake, data = msleep, geom = c("point", "smooth"))

# This is equivalent to
p6B <- ggplot(msleep, aes(sleep_rem / sleep_total, awake)) + geom_point() + geom_smooth()

ggarrange(p6A, p6B, ncol = 2, nrow = 1)
```

```{r}
p7 <- ggplot(msleep, aes(sleep_rem / sleep_total, awake))
summary(p7)
```


```{r}
p8 <- ggplot(msleep, aes(sleep_rem / sleep_total, awake)) + geom_point()
summary(p8)
```

Layers are regular R objects :a set of plots can be initialised using different data then enhanced with the same layer.

```{r, fig.height= 5, fig.width= 15}
bestfit <- geom_smooth(method = "lm", se = F, colour = alpha("steelblue", 0.5), size = 2)

p9A <- qplot(sleep_rem, sleep_total, data = msleep) + bestfit
p9B <- qplot(awake, brainwt, data = msleep, log = "y") + bestfit
p9C <- qplot(bodywt, brainwt, data = msleep, log = "xy") + bestfit

ggarrange(p9A, p9B, p9C, ncol = 3, nrow = 1)
```


Data : it must be a data frame.

```{r}
p <- ggplot(mtcars, aes(mpg, wt, colour = cyl)) + geom_point()
p
```


```{r}
mtcars2 <- transform(mtcars, mpg = mpg ^ 2)
mtcars2
```

```{r}
p %+% mtcars2
```

It is not necessary to specify a default dataset except when using faceting;
faceting is a global operation (i.e., it works on all layers) and it needs to have
a base dataset which defines the set of facets for all datasets.




own data.
The data is stored in the plot object as a copy, not a reference. This has two
important consequences: if your data changes, the plot will not; and ggplot2
objects are entirely self-contained so that they can be save()d to disk and
later load()ed and plotted without needing anything else from that session.


Aesthetic mappings

To describe the way that variables in the data are mapped to things that we
can perceive on the plot (the "aesthetics")
The aes function takes a list of aesthetic-variable pairs like these:
aes(x = weight, y = height, colour = age) : mapping x-position to weight, y-position to height and colour to age
to age.
NOTE :never refer to variables outside of the dataset (e.g., with diamonds$carat), as this makes it impossible to encapsulate all of the
data needed for plotting in a single object.

functions of variables can be used. Example : aes(weight, height, colour = sqrt(age))

Any variable in an aes() specification must be contained inside the plot or layer data.



```{r}
p451 <- ggplot(mtcars)
```

```{r}
summary(p451)
```


```{r}
p451 <- p451 + aes(wt, hp)
summary(p451)
```


```{r, fig.height= 5, fig.width= 15}
p451A <- ggplot(mtcars, aes(x = mpg, y = wt))

p451B <- p451A + geom_point() + ggtitle("default mappings in the plot p")

p451C <- p451A + geom_point(aes(colour = factor(cyl))) + ggtitle("Overriding colour with factor(cyl)")

p451D <- p451A + geom_point(aes(y = disp)) + ggtitle("overriding y-position with disp")

ggarrange(p451B, p451C, p451D, ncol = 3, nrow = 1)
```

Aesthetic mappings specified in a layer affect only that layer.


```{r, fig.width= 16, fig.height= 5}
p451E <- ggplot(data = mtcars, aes(x = mpg, y = wt)) + geom_point() + ggtitle("p451E")

p451F <- ggplot() + geom_point(data = mtcars, aes(x = mpg, y = wt)) + ggtitle("p451F")

p451G <- ggplot(data = mtcars) + geom_point(aes(x = mpg, y = wt)) + ggtitle("p451G")

p451H <- ggplot(data = mtcars, aes(x = mpg, y = wt)) + geom_point(aes(y = disp)) + ggtitle("p451H")

ggarrange(p451E, p451F,p451G ,p451H, ncol = 4, nrow = 1)
```

# in P451E, we define y = disp in aes
# in P451D, using aes of layer geom_point, the y axe is disp instead of wt (in P451A)
but the aes in the ggplot function must be defined ?

default mapping : aes (mpg, wt)
ADD : ase(colour = cyl)  result : aes(mpg, wt, colour = cyl)
Override: aes(y = disp) result : aes(mpg, disp)
Remove: aes(y= NULL) result : aes(mpg)

4.5.2 Setting vs. mapping

Aesthetics can vary for each observation being plotted, while parameters do not.
We map an aesthetic to a variable (e.g., (aes(colour = cut))) or set it to a constant (e.g.,colour = "red").

```{r, fig.width= 10}
p452 <- ggplot(mtcars, aes(mpg, wt))

p452A <- p452 + geom_point(colour = "darkblue", size = 5) + ggtitle("sets the colour of the points, \n using the colour parameter of the layer")


p452B <- p452 + geom_point(aes(colour = "darkblue"), size = 5) + ggtitle("This maps (not sets) the colour to the value \"darkblue\"")

ggarrange(p452A, p452B, ncol = 2, nrow = 1)
```

p452B : This maps (not sets) the colour to the value \darkblue". This eectively
creates a new variable containing only the value \darkblue" and then maps
colour to that new variable. Because this value is discrete, the default colour
scale uses evenly spaced colours on the colour wheel, and since there is only
one value this colour is pinkish.

4.5.3 Grouping
By default, the group is set to the interaction of all discrete variables in the plot.


```{r}
install.packages("nlme")
```

```{r}
library(nlme)
```

```{r}
str(Oxboys)
```

```{r}
head(Oxboys)
```

Multiple groups, one aesthetic.
```{r}
p453 <- ggplot(Oxboys, aes(age, height, group = Subject)) + geom_line()
```

```{r, fig.width= 15, fig.height= 5}
p453A <- ggplot(Oxboys, aes(age, height, group = Subject)) + geom_line() + ggtitle("Correctly specifying group = Subject produces one line per subject.")

p453B <- ggplot() + geom_line(data = Oxboys, aes(age, height, group = Subject)) + ggtitle("Correctly specifying group = Subject produces one line per subject.")

ggarrange(p453A, p453B, ncol = 2, nrow = 1)
```


```{r}
ggplot() + geom_point(data = Oxboys, aes(age, height))
ggplot() + geom_line(data = Oxboys, aes(age, height)) # Line plots with an incorrect grouping specification typically have this characteristic appearance.
```

Different groups on different layers.

```{r, fig.width= 15, fig.height= 5}
p453C <- p453 + geom_smooth(aes(group = Subject), 
                            method="lm", 
                            se = F) + ggtitle("Using the same grouping as the lines results in a line of best fit for each boy")
p453D <- p453 + geom_smooth(aes(group = 1), 
                            method="lm", 
                            size = 2, 
                            se = F) + ggtitle("Using aes(group = 1) in thesmooth layer fits a single line of best fit across all boys")

ggarrange(p453C, p453D, ncol = 2, nrow = 1)
```

This is not what we wanted; we have inadvertently added a smoothed line for each boy.

The new layer needs a diffierent group aesthetic, group = 1, so that the new line will be based on all the data, as shown in the second plot in
the figure.

***Overriding the default grouping***

```{r}
boysbox <- ggplot(Oxboys, aes(Occasion, height)) + geom_boxplot()
boysbox
```

```{r}
boysbox2 <- ggplot(Oxboys, aes(height,Occasion)) + geom_boxplot()
boysbox2
```
There is no need to specify the group aesthetic here; the default grouping works because occasion is a discrete variable.
```{r}
boysbox + geom_line(aes(group = Subject), colour = "#3366FF")
```

4.5.4 Matching aesthetics to graphic objects


4.6 Geoms

4.7 Stat

4.8 Position adjustments
The five position adjustments.
dodge:  Adjust position by dodging overlaps to the side
fill:  Stack overlapping objects and standardise have equal height
identity: Don't adjust position
jitter: Jitter points to avoid overplotting
stack: Stack overlapping objects on top of one another


stacking, filling and dodging :
Stacking puts bars on the same x on top of one another, filling does the same, but normalises height to 1;dodging places the bars side-by-side.

Dodging is rather similar to faceting, and the advantages and disadvantages of each method are described in Section 7.2.6.

The identity adjustment (i.e., do nothing) doesn't make much sense for bars, but is shown in Figure 4.9 along with a line plot of the same data for reference.


4.9.1 Combining geoms and stats


p491 <- ggplot(diamonds, aes(carat)) + xlim(0, 3)

p491A <- p491 + stat_bin(aes(ymax = ..count..), binwidth = 0.1, geom = "area")
p491B <- p491 + stat_bin(aes(size = ..density..), binwidth = 0.1, geom = "point", position="identity")
p491C <- p491 + stat_bin(aes(y = 1, fill = ..count..), binwidth = 0.1, geom = "tile", position="identity")




























