---
title: "Correlograms"
output: html_document
---


A correlogram or correlation matrix allows to analyse the relationship between each pair of numeric variables in a dataset. 

It gives a quick overview of the whole dataset. It is more used for exploratory purpose than explanatory.



# 1. USING THE GGALLY PACKAGE


```{r}
library(GGally)
```

The ggpairs() function of the GGally package allows to build a great scatterplot matrix.

Scatterplots of each pair of numeric variable are drawn on the left part of the figure. 

Pearson correlation is displayed on the right.

Variable distribution is available on the diagonal.



```{r}
library(tidyverse)
```

```{r}
bank <- read.csv("bank_cleaned.csv")
```

```{r}
num_bank <- bank %>% 
  select(where(is.numeric)) %>% 
  select(-c(X, response_binary))
```

```{r}
str(num_bank)
```





https://www.rdocumentation.org/packages/GGally/versions/1.5.0/topics/ggpairs

ggpairs(
  data,
  mapping = NULL,
  columns = 1:ncol(data),
  title = NULL,
  upper = list(continuous = "cor", combo = "box_no_facet", discrete = "facetbar", na = "na"),
  lower = list(continuous = "points", combo = "facethist", discrete = "facetbar", na ="na"),
  diag = list(continuous = "densityDiag", discrete = "barDiag", na = "naDiag"),
  params = NULL,
  xlab = NULL,
  ylab = NULL,
  axisLabels = c("show", "internal", "none"),
  columnLabels = colnames(data[columns]),
  labeller = "label_value",
  switch = NULL,
  showStrips = NULL,
  legend = NULL,
  cardinality_threshold = 15,
  progress = NULL,
  legends = stop("deprecated"))







```{r}
ggpairs(num_bank, title="correlogram with ggpairs()") 
```


```{r, fig.width= 10, fig.height= 7.5}
ggpairs(iris, 
        mapping = aes(color = Species),
        columns = c('Sepal.Length', 'Sepal.Width', 'Petal.Length', 'Petal.Width', 'Species'), 
        columnLabels = c('Sepal.Length', 'Sepal.Width', 'Petal.Length', 'Petal.Width', 'Species')) + 
  scale_colour_manual(values=c('red','blue','orange')) +
  theme_minimal()
```
```{r}

library(MASS)
```
```{r}
data(banknote)
```


```{r}
library(uskewFactors)
```

```{r}
data(banknote)
```

```{r}
swissbank <- banknote %>% 
  mutate(status =  case_when(
    Y == 1~ "counterfeit", 
    Y == 0 ~ "genuine"))

swissbank<- subset(swissbank, select = -c(Y))
```

This package contains measurements on 200 Swiss banknotes: 100 genuine and 100 counterfeit. The variables are length of bill, width of left edge, width of right edge , bottom margin width and top margin width. All measurements are in millimetres. The data source is noted below. This data is also available in the alr package in R.


```{r}
str(swissbank)
```



```{r, fig.height= 7.5, fig.width= 15}
ggpairs(swissbank, mapping = aes(col = status)) + scale_colour_manual(values=c('blue','orange')) +
  theme_bw()
```


## Scatterplot matrix with ggpairs()


The ggcorr() function allows to visualize the correlation of each pair of variable as a square. Note that the method argument allows to pick the correlation type you desire.

```{r}
# Nice visualization of correlations
ggcorr(num_bank, method = c("everything", "pearson")) 
```

### Split by group

```{r}
# From the help page:
data(flea)
ggpairs(flea, columns = 2:4, ggplot2::aes(colour=species)) 
```

### Change plot types
Change the type of plot used on each part of the correlogram. This is done with the upper and lower argument.

```{r}
library(reshape)
```

```{r}

data(tips)
ggpairs(tips[, c(1, 3, 4, 2)],
  upper = list(continuous = "density", combo = "box_no_facet"),
  lower = list(continuous = "points", combo = "dot_no_facet")
)
```



```{r, fig.height= 10, fig.width= 15}
ggpairs(tips,
        upper = list(continuous = "density", combo = "box_no_facet"),
        lower = list(continuous = "points", combo = "dot_no_facet"))
```




# 2. USING THE CORRGRAM PACKAGE

```{r}
library(corrgram)
```

```{r, fig.height= 7.5, fig.width= 10}
corrgram(mtcars, order = TRUE, 
         lower.panel = panel.shade, 
         upper.panel = panel.pie, 
         text.panel = panel.txt, 
         main = "Car Milage Data in PC2/PC1 Order") 
```
```{r, fig.height= 7.5, fig.width= 10}
corrgram(mtcars, 
         order = TRUE, 
         lower.panel = panel.ellipse, 
         upper.panel = panel.pts, 
         text.panel = panel.txt, 
         diag.panel = panel.minmax, 
         main="Car Milage Data in PC2/PC1 Order") 
```

```{r, fig.height= 7.5, fig.width= 10}
corrgram(mtcars, order = NULL, 
         lower.panel = panel.shade, 
         upper.panel = NULL, 
         text.panel = panel.txt, 
         main="Car Milage Data (unsorted)")
```

# 3. OTHER METHODS

Lesser known ways to build correlogram with R, like the ellipse package, the plot() function and the car package.

## Package ellipse

```{r}
library(ellipse)
```
```{r}
library(RColorBrewer)
```

Use of the mtcars data proposed by R

```{r}
data <- cor(mtcars)
```

```{r}
# Build a Pannel of 100 colors with RcolorBrewer
my_colors <- brewer.pal(5, "Spectral")
my_colors <- colorRampPalette(my_colors)(100)
 
# Order the correlation matrix
ord <- order(data[1, ])

data_ord <- data[ord, ord]

plotcorr(data_ord , col=my_colors[data_ord*50+50] , mar=c(1,1,1,1)  )
```

The native plot() function does the job pretty well as long as you just need to display scatterplots.


## The native plot() function


```{r, fig.height= 7.5, fig.width= 10}
# Plot
plot(mtcars , pch=20 , cex=1.5 , col="blue")
```

























