---
title: "ggplot2-Toolbox"
author: "Minh Trung DANG"
date: "24/05/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(ggplot2)
library(ggpubr)
```

5.3 Basic plot types
```{r}
df <- data.frame(
x = c(3, 1, 5),
y = c(2, 4, 6),
label = c("a","b","c")
)
```

```{r}
df
```

```{r}
p <- ggplot(df, aes(x, y, label = label)) + xlab(NULL) + ylab(NULL)
p
```


```{r}
p + geom_point()
```

```{r}
p + geom_bar(stat="identity") + ggtitle("geom_bar(stat=\"identity\")")
```

```{r}
p + geom_line() + ggtitle("geom_line")
```
```{r}
p + geom_area() + ggtitle("geom_area")
```
```{r}
p + geom_path() + ggtitle("geom_path")
```
```{r}
p + geom_text() + ggtitle("geom_text")
```
```{r}
p + geom_tile() + ggtitle("geom_tile")
```
```{r}
p + geom_polygon() + ggtitle("geom_polygon")
```

5.4 Displaying distributions

1d continuous distributions :  histogram

compare the distribution between groups: 
option 1: create small multiples of the histogram, facets = . ~ var; use a frequency polygon, geom = "freqpoly"
option 2: create a conditional density plot, position = "fill".


```{r}
depth_dist <- ggplot(diamonds, aes(depth)) + xlim(58, 68)
```

```{r}
depth_dist + geom_histogram(aes(y = ..density..), binwidth = 0.1) + facet_grid(cut ~ .) + ggtitle("faceted histogram")
```

```{r}
depth_dist + geom_histogram(aes(fill = cut), binwidth = 0.1, position = "fill") + ggtitle("conditional density plot")
```

```{r}
depth_dist + geom_freqpoly(aes(y = ..density.., colour = cut), binwidth = 0.1) + ggtitle("frequency polygons")
```

Both the histogram and frequency polygon geom use stat_bin.This statistic produces two output variables count and density.
(The count is the default as it is most interpretable)


The density is basically the count divided by the total count, and is useful when you want to compare the shape of the distributions,
not the overall size. You will often prefer this when comparing the distribution
of subsets that have different sizes.

All show an interesting pattern: as quality increases, the distribution shifts to the left and becomes more symmetric.


A continuous variable conditioned by a categorical variable
geom_boxplot = stat_boxplot + geom_boxplot: box-and-whisker plot
This is a useful display when the categorical variable has many distinct values.
When there are few values, the techniques described above give a better view of the shape of the distribution.
This technique can also be used for continuous variables, if they are first finely binned.

```{r}
library(plyr)
```

```{r, fig.height =  5, fig.width= 15}
p54A <- qplot(cut, depth, data=diamonds, geom="boxplot") + ggtitle("boxplots conditioned on categorical variable")

p54B <- qplot(carat, depth, data=diamonds, geom="boxplot", group = round_any(carat, 0.1, floor), xlim = c(0, 4)) + ggtitle("boxplots conditioned on continuous variable")

# group = round any(carat, 0.1, floor) is used to get a boxplot for each 0.1 carat bin.
ggarrange(p54A, p54B, ncol = 2, nrow = 1)
```

geom_jitter = position_jitter + geom_point: a crude way of looking at discrete distributions by adding random noise to the discrete values so that they don't overplot.


```{r, fig.height =  5, fig.width= 15}
p54C <- qplot(class, cty, data=mpg, geom="jitter")
p54D <- qplot(class, drv, data=mpg, geom="jitter")

ggarrange(p54C, p54D, ncol = 2, nrow = 1)
```


geom_density = stat_density + geom_area: a smoothed version of the frequency polygon based on kernel smoothers. 
Use a density plot when you know that the underlying density is smooth, continuous and unbounded.
You can use the adjust parameter to make the density more or less smooth.

```{r}
p54E <-qplot(depth, data=diamonds, geom="density", xlim = c(54, 70))
p54F <-qplot(depth, data=diamonds, geom="density", xlim = c(54, 70), fill = cut, alpha = I(0.2))

ggarrange(p54E, p54F, ncol = 2, nrow = 1)
```

5.5 Dealing with overplotting

A. Making the points smaller
Small amounts of overplotting can sometimes be alleviated by making the points smaller, or using hollow glyphs.

```{r, fig.height= 8, fig.width= 8}
df55 <- data.frame(x = rnorm(2000), y = rnorm(2000))
 p55A <-  ggplot(df55, aes(x, y))
 p55B <-  p55A  + geom_point()
 p55C <-  p55A + geom_point(shape = 1)
 p55D <-  p55A  + geom_point(shape = ".") # Pixel sized

ggarrange(p55A, p55B, p55C, p55D, ncol = 2, nrow = 2)
```

B. Using alpha blending (transparency) make the points transparent.
If you specify alpha as a ratio, the denominator gives the number of points that must be overplotted
to give a solid colour

```{r, fig.height= 4, fig.width= 10}
p55E <- p55A + geom_point(colour = alpha("black", 1/3))
p55F <- p55A + geom_point(colour = alpha("black", 1/5))
p55G <- p55A + geom_point(colour = alpha("black", 1/10))

ggarrange(p55E, p55F, p55G, ncol = 3, nrow = 1)
```

C. randomly jitter the point to alleviate some overlaps.

This is particularly useful in conjunction with transparency.
 
```{r}
p55H <- ggplot(diamonds, aes(table, depth)) + xlim(50, 70) + ylim(50, 70)
p55I <- p55H + geom_point()
p55J <- p55H + geom_jitter()

ggarrange(p55I, p55J, ncol = 2, nrow = 1)
```
```{r}
jit <- position_jitter(width = 0.5)
p55K <- p55H  + geom_jitter(position = jit)
p55L <- p55H  + geom_jitter(position = jit, colour = alpha("black", 1/10))
p55M <- p55H  + geom_jitter(position = jit, colour = alpha("black", 1/50))
p55N <- p55H  + geom_jitter(position = jit, colour = alpha("black", 1/200))


ggarrange(p55K, p55L,p55M, p55N,  ncol = 2, nrow = 2)
```








