---
title: "ggplot2-second edition-chapter2"
date: "24/05/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
head(mpg)
```

• cty and hwy record miles per gallon (mpg) for city and highway driving.
• displ is the engine displacement in litres.
• drv is the drivetrain: front wheel (f), rear wheel (r) or four wheel (4).
• model is the model of car. There are 38 models, selected because they had
a new edition every year between 1999 and 2008.
• class (not shown), is a categorical variable describing the “type” of car: two seater, SUV, compact, etc.

Every ggplot2 plot has three key components:
1. data
2. A set of aesthetic mappings between variables in the data and visual properties
3. At least one layer which describes how to render each observation. Layers are usually created with a geom function.

```{r}
library(ggplot2)
library(ggpubr)
```


```{r}
ggplot(mpg, aes(x = displ, y = hwy)) + geom_point()
```


This produces a scatterplot defined by:
1. Data: mpg.
2. Aesthetic mapping: engine size mapped to x position, fuel economy to y position.
3. Layer: points.

Note: data and aesthetic mappings are supplied in ggplot(), then layers are added on with +.


The plot shows a strong correlation: as the engine size gets bigger, the fuel economy gets worse
There are also some interesting outliers: some cars with large engines get higher fuel economy than average

Exercises
1. How would you describe the relationship between cty and hwy? Do you have any concerns about drawing conclusions from that plot?

```{r}
ggplot(mpg, aes(x = cty, y = hwy)) + geom_point()
```

2. What does ggplot(mpg, aes(model, manufacturer)) + geom point() show? 
Is it useful? How could you modify the data to make it more informative?

```{r, fig.width= 10}
ggplot(mpg, aes(model,manufacturer)) + geom_point() + coord_flip()
```

one manufaturer can creat many car model. To make the graphe more informative, we can regroup all the models for each manufaturer (order?)

Colour, Size, Shape and Other Aesthetic Attributes


There is one scale for each aesthetic mapping in a plot.
The scale is also responsible for creating a guide, an axis or legend, that allows you to read the plot, converting aesthetic values back into data values

Mapping the class variable to color:

```{r, fig.width= 5, fig.height= 5}
ggplot(mpg, aes(displ, cty, colour = class)) + 
  geom_point()
```
Each point a unique colour corresponding to its class
The group of cars with unusually high fuel economy for their engine size are two seaters: cars with big engines, but lightweight bodies.


If you want to set an aesthetic to a fixed value, without scaling it, do so in the individual layer outside of aes().

```{r}

p241A <- ggplot(mpg, aes(displ, hwy)) + geom_point(aes(colour = "blue"))

p241B <- ggplot(mpg, aes(displ, hwy)) + geom_point(colour = "blue") # outside of aes()

ggarrange(p241A, p241B, ncol= 2, nrow =1)
```


2.5 Facetting
Facetting creates tables of graphics by splitting the data into subsets and displaying the same graph for each subset.


```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point() +
  facet_wrap(~class)
```
when to use facetting and when to use aesthetics ?

```{r}
str(mpg)
```

```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point() +
  facet_wrap(~cyl)
```


2.6 Adding a Smoother to a Plot

```{r}
ggplot(mpg, aes(displ, hwy)) + 
  geom_point() +
  geom_smooth()
```

Important : "method"


• method = "loess", the default for small n, uses a smooth local regression
(as described in ?loess). The wiggliness of the line is controlled by the span
parameter, which ranges from 0 (exceedingly wiggly) to 1 (not so wiggly).


```{r}
p25A <- ggplot(mpg, aes(displ, hwy)) +
geom_point() +
geom_smooth(span = 0.2)

p25B <- ggplot(mpg, aes(displ, hwy)) +
geom_point() +
geom_smooth(span = 1)


ggarrange(p25A, p25B, ncol = 2, nrow =1)
```

NOTE : Loess does not work well for large datasets (it’s O(n2) in memory), so an
alternative smoothing algorithm is used when n is greater than 1000

• method = "gam" fits a generalised additive model provided by the mgcv
package. You need to first load mgcv, then use a formula like formula = y
˜ s(x) or y ˜ s(x, bs = "cs") (for large data). This is what ggplot2 uses
when there are more than 1000 points.

```{r}
library(mgcv)
```

```{r}
ggplot(mpg, aes(displ, hwy)) +
geom_point() +
geom_smooth(method = "gam", formula = y ~ s(x))
```

• method = "lm" fits a linear model, giving the line of best fit.

```{r}
ggplot(mpg, aes(displ, hwy)) + geom_point() +
  geom_smooth(method = "lm")
```

• method = "rlm" works like lm(), but uses a robust fitting algorithm so that
outliers don’t affect the fit as much.


2.6.2 Boxplots and Jittered Points
to know how the values of the continuous variables vary with the levels of the categorical variable


how fuel economy varies within car class


```{r, fig.height= 3, fig.width= 3}
ggplot(mpg, aes(drv, hwy)) + 
  geom_point()
```
• Jittering, geom jitter(), adds a little random noise to the data which can
help avoid overplotting.
• Boxplots, geom boxplot(), summarise the shape of the distribution with a
handful of summary statistics.
• Violin plots, geom violin(), show a compact representation of the “density”
of the distribution, highlighting the areas where more points are found.

```{r, fig.height= 5, fig.width= 12}
p262A <- ggplot(mpg, aes(drv, hwy)) + geom_jitter()
p262B <- ggplot(mpg, aes(drv, hwy)) + geom_boxplot()
p262C <- ggplot(mpg, aes(drv, hwy)) + geom_violin()

ggarrange(p262A,p262B, p262C, ncol =3, nrow =1)
```

Boxplots summarise the bulk of the distribution with only five numbers, while jittered plots show
every point but only work with relatively small datasets.
Violin plots give the richest display, but rely on the calculation of a density estimate, which
can be hard to interpret.

For jittered points, geom jitter() offers the same control over aesthetics as geom point(): size, colour, and shape. 
For geom boxplot() and geom violin(), you can control the outline colour or the internal fill colour


2.6.3 Histograms and Frequency Polygons


```{r}

p263A <- ggplot(mpg, aes(hwy)) + geom_histogram()
# stat_bin() using bins = 30. Pick better value with binwidth.

p263B <- ggplot(mpg, aes(hwy)) + geom_freqpoly()
#> stat_bin() using bins = 30. Pick better value with binwidth

ggarrange(p263A, p263B, ncol =2, nrow =1)
```

Both histograms and frequency polygons work in the same way: 
* bin the data
* count the number of observations in each bin

The only difference: histograms use bars and frequency polygons use lines.


```{r}
p263C <- ggplot(mpg, aes(hwy)) + geom_freqpoly(binwidth = 2.5)

p263D <- ggplot(mpg, aes(hwy)) + geom_freqpoly(binwidth = 1)

ggarrange(p263C, p263D, ncol = 2, nrow = 1)
```


To compare the distributions of different subgroups, you can map a categorical variable to either fill (for geom histogram()) or colour (for geom freqpoly()).


It’s easier to compare distributions using the frequency polygon because the underlying perceptual task is easier. You can also use facetting: this makes comparisons a little harder, but it’s easier to see the distribution of each group.

```{r}
p263E <- ggplot(mpg, aes(displ, colour = drv)) + 
  geom_freqpoly(binwidth = 0.5)

p263F <- ggplot(mpg, aes(displ, fill = drv)) + 
  geom_histogram(binwidth = 0.5) + 
  facet_wrap(~drv, ncol = 1)

ggarrange(p263E, p263F, ncol = 2, nrow = 1)
```


2.6.4 Bar Charts

```{r, fig.height= 5, fig.width= 10}
ggplot(mpg, aes(manufacturer)) +
  geom_bar()
```

```{r}
drugs <- data.frame(
  drug = c("a", "b", "c"),
  effect = c(4.2, 9.7, 6.1))
```

```{r}
p254A <- ggplot(drugs, aes(drug, effect)) + geom_bar()
p254A
```

To display this sort of data, you need to tell geom bar() to not run the
default stat which bins and counts the data.


```{r}
p254B <- ggplot(drugs, aes(drug, effect)) + geom_bar(stat = "identity")
p254B
```

Solution : it is better to use geom_point() because points take  up less space than bars, and do not require that the y axis inlcudes 0


```{r}
ggplot(drugs, aes(drug, effect)) + geom_point()
```

2.6.5 Time Series with Line and Path Plots

typically used for time series data.
Line plots : join the point from left to right
              usually have time on the x-axis, showing how a single varauble has changed over time.

Path plots : join the points in the order that they appear in the dataset
             show how two variables have simultaneously changed over time, with time encoded in the way that
             observations are connected.

```{r}
p265A <- ggplot(economics, aes(date, unemploy/pop)) + geom_line()

p265B <-ggplot(economics, aes(date, uempmed)) + geom_line()

ggarrange(p265A, p265B, ncol= 2, nrow = 1)

```

```{r}
year <- function(x) as.POSIXlt(x)$year + 1900
```
)

```{r, fig.height= 5, fig.width= 12.5}
p265C <- ggplot(economics, aes(unemploy/pop, uempmed)) + 
  geom_path() +
  geom_point()

p265D <- ggplot(economics, aes(unemploy/pop, uempmed)) +
  geom_path(colour = "grey50") +
  geom_point(aes(colour = year(date)))

ggarrange(p265C, p265D, ncol= 2, nrow = 1)

```

2.7 Modifying the Axes

```{r, fig.height = 5, fig.width= 15}
p27A <- ggplot(mpg, aes(cty, hwy)) +
  geom_point(alpha = 1/3)

p27B <- ggplot(mpg, aes(cty, hwy)) + geom_point(alpha = 1/3) +
  xlab("city driving (mpg)") +
  ylab("highway driving (mpg)")

p27C <- ggplot(mpg, aes(cty, hwy)) +
  geom_point(alpha = 1/3) +
  xlab(NULL) +
  ylab(NULL)

ggarrange(p27A, p27B, p27C, ncol= 3, nrow = 1)
```

```{r, fig.height = 5, fig.width= 15}
p27D <- ggplot(mpg, aes(drv, hwy)) +
  geom_jitter(width = 0.25)

p27E <- ggplot(mpg, aes(drv, hwy)) +
  geom_jitter(width = 0.25) +
  xlim("f", "r") +
  ylim(20, 30)
# Warning: Removed 138 rows containing missing values (geom_point).
# For continuous scales, use NA to set only one limit

p27F <- ggplot(mpg, aes(drv, hwy)) +
  geom_jitter(width = 0.25, na.rm = TRUE) +
  ylim(NA, 30)

ggarrange(p27D, p27E, p27F, ncol= 3, nrow = 1)
```







































