---
title: "Notes for gglot2"
author: "Minh Trung DANG"
date: "23/05/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



```{r cars}
library(ggplot2)
```

```{r}
head(diamonds)
```

```{r}
dsmall <- diamonds[sample(nrow(diamonds), 100), ]
```

```{r}
dsmall2 <- diamonds[sample(nrow(diamonds), 8000), ]
```


```{r}
qplot(carat, price, data = diamonds)
```

The plot shows a strong correlation with notable outliers and some interest-
ing vertical striation. The relationship looks exponential, though, so the first
thing we'd like to do is to transform the variables. Because qplot() accepts
functions of variables as arguments, we plot log(price) vs. log(carat):

```{r}
qplot(log(carat), log(price), data = diamonds)
```

Arguments can also be combinations of existing variables, so, if we are curi-
ous about the relationship between the volume of the diamond (approximated
by x * y * z) and its weight, we could do the following:

```{r}
qplot(carat, x * y * z, data = diamonds)
```

```{r}
str(diamonds)
```

Colour, size, shape and other aesthetic attributes

```{r}
qplot(carat, price, data = diamonds, colour = color)
```

```{r}
qplot(carat, price, data = diamonds, shape = cut)
```


```{r}
qplot(carat, price, data = diamonds, alpha = I(1/10))
qplot(carat, price, data = diamonds, alpha = I(1/100))
qplot(carat, price, data = diamonds, alpha = I(1/200))
```

```{r}
qplot(carat, price, data = diamonds, color = I("red"))
```


```{r}
qplot(carat, price, data = diamonds, 
      color = I("blue"), 
      alpha = I(0.01))
```

```{r}
qplot(carat, price, data = diamonds, 
      color = c("blue"), 
      alpha = 0.01 )
```
This above code does not work.

Plot geom
geom = geometric object

Two-dimensional relationships
***geom = "point"*** draws points to produce a scatterplot. 
This is the default when you supply both x and y arguments to qplot().

***geom = "smooth"*** fits a smoother to the data and displays the smooth and its standard error
***geom = "boxplot" *** produces a box-and-whisker plot to summarise the distribution of a set of points, x 2.5.2.
***geom = "path" and geom = "line" *** draw lines between the data points.
Traditionally these are used to explore relationships between time and another variable, but lines may be used to join observations connected in some other way. 
A line plot is constrained to produce lines that travel from left to right, while paths can go in any direction,

One-dimensional relationships

***For continuous variables***
geom = "histogram" draws a histogram 
geom = "freqpoly" a frequency polygon
geom = "density" creates a density plot
The histogram geom is the default when you only supply an x value to qplot().

***For discrete variables**
geom = "bar" makes a bar chart,


Adding a smoother to a plot
to see exactly what trend is shown by the data


```{r}
qplot(carat, price, data = diamonds, geom = c("point", "smooth"))
```



If you want to turn the confidence interval oof, use se = FALSE.

```{r}
qplot(carat, price, data = dsmall, 
      geom = c("point"))
```


```{r}
qplot(carat, price, data = dsmall, 
      geom = c("point", "smooth"))
```

```{r}
qplot(carat, price, data = dsmall, 
      geom = c("point", "smooth"), 
      se = FALSE)
```

```{r}
qplot(carat, price, data = dsmall2,
      colour = color,
      geom = c("point", "smooth"), 
      se = FALSE)
```


```{r}
qplot(carat, price, data = dsmall2,
      colour = color,
      colour = I("royalblue"),
      geom = c("point", "smooth"), 
      se = FALSE)
```
# the argument color =  colour is ignored.




method = "loess", the default for small n, uses a smooth local regression.
More details about the algorithm used can be found in ?loess. The
wiggliness of the line is controlled by the span parameter, which ranges
from 0 (exceedingly wiggly) to 1 (not so wiggly), as shown in Figure 2.5.

```{r}
qplot(carat, price, data = dsmall, geom = c("point", "smooth"),
span = 0.2)
```
```{r}
qplot(carat, price, data = dsmall, geom = c("point", "smooth"), span = 1)
```
!!! Loess does not work well for large datasets

method = "lm" fitts a linear model. The default will fit a straight line to
your data, or you can specify formula = y ~ poly(x, 2) to specify a
degree 2 polynomial, or better, load the splines package and use a natural
spline: formula = y ~ ns(x, 2). The second parameter is the degrees
of freedom: a higher number will create a wigglier curve. You are free to
specify any formula involving x and y.

```{r}
library(splines)
library(ggplot2)
```

```{r}
qplot(carat, price, data = dsmall, 
      geom = c("point", "smooth"),
      method = "lm")
```

```{r}
qplot(carat, price, data = dsmall, 
      geom = c("point", "smooth"),
      method = "lm",
      formula = y ~ ns(x,5))
```

Boxplots and jittered points
to know how the values of the continuous variables vary with the levels of the categorical variable
***jittering (geom = "jitter") 
***box-and-whisker plots (geom ="boxplot").



```{r}
qplot(color, price, data = dsmall2, 
      geom = "jitter")
```


```{r}
qplot(color, price, data = dsmall2, 
      geom = "boxplot")
```

Boxplots summarise the bulk of the distribution with only five numbers, while jittered plots show every
point but can suffer from overplotting.

(there is very little change in the median and adjacent quartiles)

To impprove jitter plot
```{r}
qplot(color, price, data = diamonds, 
      geom = "jitter", 
      alpha = I(1/50))
```
```{r}
qplot(color, price/carat, data = diamonds, 
      geom = "jitter", 
      alpha = I(1/50))
```

This technique can't show the positions of the quantiles as well as a boxplot can, but it may reveal other features of the distribution that a boxplot cannot.


For jittered points, qplot offers the same control over aesthetics as it
does for a normal scatterplot: size, colour and shape. 
For boxplots you can control the outline colour, the internal fill colour and the size of the lines.


```{r}
qplot(color, price/carat, data = diamonds,
      colour = color,
      geom = "jitter", 
      alpha = I(1/50))
```
```{r}
qplot(color, price/carat, data = diamonds,
      shape = color,
      geom = "jitter", 
      alpha = I(1/50))
```


```{r}
qplot(color, price, data = dsmall2, 
      fill = color,
      geom = "boxplot")
```

```{r}
qplot(color, price, data = dsmall2,
      colour = c("red"),
      fill = color,
      geom = "boxplot")
```
# this above code does not work

Another way to look at conditional distributions is to use faceting to plot
a separate histogram or density plot for each value of the categorical variable.


Histogram and density plots
To show the distribution of a single variable.


```{r}
qplot(carat, data = diamonds, 
      geom = "histogram")
```

```{r}
qplot(carat, data = diamonds, 
      geom = "density")
```


```{r}
qplot(price, data = diamonds, 
      geom = "histogram")
```

```{r}
qplot(price, data = diamonds, 
      geom = "density")
```

```{r}
qplot(x*y*z, data = diamonds, 
      geom = "histogram")
```

```{r}
qplot(x*y*z, data = diamonds, 
      geom = "density")
```

Playimg with binwithd
```{r}
qplot(carat, data = diamonds, 
      geom = "histogram",
      binwidth = 1,
      xlim = c(0,3))


qplot(carat, data = diamonds, 
      geom = "histogram", 
      binwidth = 0.1,
      xlim = c(0,3))

qplot(carat, data = diamonds, 
      geom = "histogram", 
      binwidth = 0.01,
      xlim = c(0,3))
```

To compare the distributions of different subgroups, just add an aesthetic mapping, as in the following code.

```{r}
library(ggpubr)
```

```{r, fig.height= 5, fig.width=10}
p1A <- qplot(carat, data = diamonds, 
      geom = "density", 
      colour = color)

p1B <-qplot(carat, data = diamonds, 
      geom = "histogram", 
      fill = color)

ggarrange(p1A, p1B, ncol = 2, nrow = 1)
```

Bar charts
The bar geom counts the number of instances of each class so that you don't need to
tabulate your values beforehand.



If the data has already been tabulated or if you'd like to tabulate class members in some other
way, such as by summing up a continuous variable, you can use the weight geom.

```{r, fig.height= 5, fig.width= 12.5}
p2A <- qplot(color, data = diamonds, geom = "bar")

p2B <- qplot(color, data = diamonds, geom = "bar",
             fill = color)

p2C <- qplot(color, data = diamonds, geom = "bar",
             fill = cut)

ggarrange(p2A, p2B, p2C, ncol = 3, nrow = 1)
```


```{r}
qplot(color, data = diamonds, geom = "bar", 
      weight = carat)+ # the plot is weighted by weight = carat to show the total weight of diamonds of each colour. + 
scale_y_continuous("volume")
```

***Time series with line and path plots***
typically used for time series data.
Line plots join the points from left to right, while path plots join them in the order that they appear in the dataset (a line plot is just a path plot of the data sorted by x value)
Line plots usually have time on the x-axis, showing how a single variable has changed over time.
Path plots show how two variables have simultaneously changed over time, with time encoded in the way that the points are joined together.

```{r}
str(economics)
```

```{r, Percent of population that is unemployed}
qplot(date, unemploy/pop, data = economics, 
      geom = "line") +
  ggtitle("Percent of population that is unemployed")
```
```{r}
qplot(date, unemploy/pop, data = economics, 
      geom = "line") +
  ggtitle("Percent of population that is unemployed")
```


```{r}
library(ggplot2)
```
```{r}
year <- function(x) as.POSIXlt(x)$year + 1900
```

```{r, fig.width= 15, fig.height = 10}

p3A <- qplot(unemploy/pop, uempmed, data = economics,
      geom = c("point")) 

p3B <- qplot(unemploy/pop, uempmed, data = economics,
      geom = c("point", "path")) 


p3C <- qplot(unemploy / pop, uempmed, data = economics,
             geom = c("path"),  # or geom = "path"
             colour = year(date)) # apply the colour aesthetic to the line to make it easier to see the direction of time.

p3D <- qplot(unemploy / pop, uempmed, data = economics,
             geom = c("point", "path"), 
             colour = year(date))  # apply the colour aesthetic to the line to make it easier to see the direction of time.


ggarrange(p3A, p3B,p3C, p3D, ncol =2, nrow = 2)
```


Faceting
It creates tables of graphics by splitting the data into subsets and displaying the same graph for each subset in an arrangement that facilitates comparison.



```{r, fig.width= 10, fig.height= 10}
p4A <- qplot(carat, data = diamonds, 
      facets = color ~ .,
      geom = "histogram", 
      binwidth = 0.1, 
      xlim = c(0, 3))


p4B <- qplot(carat, ..density..,
      data = diamonds, 
      facets = color ~ .,
      geom = "histogram", 
      binwidth = 0.1, 
      xlim = c(0, 3))

ggarrange(p4A, p4B, ncol = 2, nrow = 1)

```

***Other options***
These are a few other qplot options to control the graphic's appearance. 
These all have the same effect as their plot equivalents:
*** xlim, ylim: set limits for the x- and y-axes, each a numeric vector of length two, 
    e.g., xlim=c(0, 20) or ylim=c(-0.9, -0.5).
    
****log: a character vector indicating which (if any) axes should be logged.
    For example, log="x" will log the x-axis, log="xy" will log both.

*** main: main title for the plot, centered in large text at the top of the plot.
This can be a string (e.g., main="plot title") or an expression 
(e.g., main = expression(beta[1] == 1)). See ?plotmath for more examples of using mathematical formulae.

***xlab, ylab: labels for the x- and y-axes. As with the plot title, these can
be character strings or mathematical expressions.







