---
title: "ggplot2-second edition-chapter3"
author: "Minh Trung DANG"
date: "24/05/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
library(ggplot2)
library(ggpubr)
```


3.3 Labels
geom text()

```{r}
df <- data.frame(x = 1, y = 3:1, family = c("sans", "serif", "mono"))
df
```


```{r}
ggplot(df, aes(x, y)) +
  geom_text(aes(label = family, family = family))
```


• fontface specifies the face: “plain” (the default), “bold” or “italic”.


df33 <- data.frame(x = 1, y = 3:1, face = c("plain", "bold", "italic"))
p33A <- ggplot(df, aes(x, y)) +
  geom_text(aes(label = face, fontface = "face"))


You can adjust the alignment of the text with the hjust (“left”, “center”,
“right”, “inward”, “outward”) and vjust (“bottom”, “middle”, “top”, “inward”,
“outward”) aesthetics. The default alignment is centered. One of
the most useful alignments is “inward”: it aligns text towards the middle
of the plot:



```{r}
df33A <- data.frame(x = c(1, 1, 2, 2, 1.5),
                    y = c(1, 2, 1, 2, 1.5),
                    text = c("bottom-left", "bottom-right","top-left", "top-right", "center"))
df33A
```


```{r}
p33B <- ggplot(df33A, aes(x, y)) + geom_text(aes(label = text))
p33C <- ggplot(df33A, aes(x, y)) + geom_text(aes(label = text), vjust = "inward", hjust = "inward")

ggarrange(p33B, p33C, ncol = 2, nrow = 1)
```

• size controls the font size. Unlike most tools, ggplot2 uses mm, rather
than the usual points (pts). This makes it consistent with other size units
in ggplot2.
• angle specifies the rotation of the text in degrees.

geom text() also has three parameters. Unlike the aesthetics, these only take
single values, so they must be the same for all labels


Often you want to label existing points on the plot. 
You don’t want the text to overlap with the points (or bars etc), so it’s useful to offset the text a little. 
The nudge x and nudge y parameters allow you to nudge the text a little horizontally or vertically:

```{r}
df33B <- data.frame(trt = c("a", "b", "c"), resp = c(1.2, 3.4, 2.5))
df33B
```


```{r, fig.width= 12, fig.height=4}
p33D <- ggplot(df33B, aes(resp, trt)) + 
  geom_point() +
  xlim(1, 3.6)

p33E <- ggplot(df33B, aes(resp, trt)) + 
  geom_point() +
  geom_text(aes(label = paste0("(", resp, ")"))) +
  xlim(1, 3.6)

p33F <- ggplot(df33B, aes(resp, trt)) + 
  geom_point() +
  geom_text(aes(label = paste0("(", resp, ")")), 
            nudge_y = -0.25) +
  xlim(1, 3.6)

ggarrange(p33D, p33E, p33F, ncol = 3, nrow =1 )
```

• If check overlap = TRUE, overlapping labels will be automatically removed.
The algorithm is simple: labels are plotted in the order they appear in the
data frame; if a label would overlap with an existing point, it’s omitted.
This is not incredibly useful, but can be handy.

```{r, fig.width = 10, fig.height = 5}
p33G <- ggplot(mpg, aes(displ, hwy)) + 
  geom_text(aes(label = model)) + 
  xlim(1, 8)

p33H <- ggplot(mpg, aes(displ, hwy)) + 
  geom_text(aes(label = model), 
            check_overlap = TRUE) 
+ xlim(1, 8)

ggarrange(p33G, p33H,  ncol = 2, nrow = 1)
```

A variation on geom text() is geom label(): it draws a rounded rectangle behind the text. 
This makes it useful for adding labels to plots with busy backgrounds:

```{r}
label <- data.frame(waiting = c(55, 80),
                    eruptions = c(2, 4.3),
                    label = c("peak one", "peak two"))
label 
```

```{r}
head(faithfuld)
```


```{r, fig.width = 10, fig.height = 5}

p33I <- ggplot(faithfuld, aes(waiting, eruptions)) +
  geom_tile(aes(fill = density)) 

p33J <- ggplot(faithfuld, aes(waiting, eruptions)) +
  geom_tile(aes(fill = density)) +
  geom_label(data = label, aes(label = label))

ggarrange(p33I, p33J,  ncol = 2, nrow = 1)
```


since a label has an absolute size (e.g. 3 cm), regardless of the size of the plot
Instead, you’ll need to tweak xlim() and ylim() based on your data and plot size.

```{r, fig.width = 10, fig.height = 5}
p33J <- ggplot(mpg, aes(displ, hwy, colour = class)) +
  geom_point()

p33K <-ggplot(mpg, aes(displ, hwy, colour = class)) +
  geom_point(show.legend = FALSE) +
  directlabels::geom_dl(aes(label = class), method = "smart.grid")

ggarrange(p33J, p33K,  ncol = 2, nrow = 1)
```

3.4 Annotations
Annotations add metadata to your plot. But metadata is just data, so you can use:
• geom_text() to add text descriptions or to label points Most plots will
not benefit from adding text to every single observation on the plot, but
labelling outliers and other important points is very useful.
• geom_rect() to highlight interesting rectangular regions of the plot.
geom_rect() has aesthetics xmin, xmax, ymin and ymax.
• geom line(), geom path() and geom segment() to add lines. All these geoms
have an arrow parameter, which allows you to place an arrowhead on the
line. Create arrowheads with arrow(), which has arguments angle, length,
ends and type.
• geom vline(), geom hline() and geom abline() allow you to add reference
lines (sometimes called rules), that span the full range of the plot.

Typically, you can either put annotations in the foreground (using alpha if
needed so you can still see the data), or in the background. With the default
background, a thick white line makes a useful reference: it’s easy to see but
it doesn’t jump out at you.

```{r}
ggplot(economics, aes(date, unemploy)) +
geom_line()
```



We can annotate this plot with which president was in power at the time.
There is little new in this code - it’s a straightforwardmanipulation of existing
geoms. There is one special thing to note: the use of -Inf and Inf as positions.
These refer to the top and bottom (or left and right) limits of the plot.

```{r}
presidential <- subset(presidential, start > economics$date[1])
presidential 
```

```{r}
ggplot(economics) +
  geom_rect(aes(xmin = start, xmax = end, fill = party),
            ymin = -Inf, ymax = Inf, alpha = 0.2,
            data = presidential) +
  geom_vline(aes(xintercept = as.numeric(start)),
             data = presidential,
             colour = "grey50", alpha = 0.5) +
  geom_text(aes(x = start, y = 2500, label = name),
            data = presidential,
            size = 3, vjust = 0, hjust = 0, nudge_x = 50) +
  geom_line(aes(date, unemploy)) +
  scale_fill_manual(values = c("blue", "red"))
```


```{r, fig.height= 10, fig.width= 20}
p33L <- ggplot(economics) +
  geom_rect(aes(xmin = start, xmax = end, fill = party),
            ymin = -Inf, ymax = Inf, alpha = 0.2,
            data = presidential) +
  scale_fill_manual(values = c("blue", "red"))

p33M <- p33L +
  geom_vline(aes(xintercept = as.numeric(start)),
             data = presidential,
             colour = "grey50", alpha = 0.5) +
  scale_fill_manual(values = c("blue", "red"))

p33N <- p33L + 
  geom_vline(aes(xintercept = as.numeric(start)),
             data = presidential,
             colour = "grey50", alpha = 0.5) +
  geom_text(aes(x = start, y = 2500, label = name),
            data = presidential,
            size = 3, vjust = 0, hjust = 0, nudge_x = 50) +
  scale_fill_manual(values = c("blue", "red"))

p33O <- p33L +
  geom_vline(aes(xintercept = as.numeric(start)),
             data = presidential,
             colour = "grey50", alpha = 0.5) +
  geom_text(aes(x = start, y = 2500, label = name),
            data = presidential,
            size = 3, vjust = 0, hjust = 0, nudge_x = 50) +
  geom_line(aes(date, unemploy)) +
  scale_fill_manual(values = c("blue", "red"))

ggarrange(p33L, p33M,p33N, p33O,  ncol = 2, nrow = 2)
```



```{r, fig.height= 5, fig.width= 20}
ggplot(diamonds, aes(log10(carat), log10(price))) +
  geom_bin2d() +
  facet_wrap(~cut, nrow = 1)
```

```{r}
mod_coef <- coef(lm(log10(price) ~ log10(carat), data = diamonds))
mod_coef
```
```{r}
mod_coef[1]
```
```{r}
mod_coef[2]
```

```{r, fig.height= 5, fig.width= 20}
ggplot(diamonds, aes(log10(carat), log10(price))) + 
  geom_bin2d() +
  geom_abline(intercept = mod_coef[1], slope = mod_coef[2],colour = "white", size = 1) +
  facet_wrap(~ cut, nrow = 1)
```





















