---
title: "Visualization Proportion-Stacked bars"
author: "Minh Trung DANG"
date: "11/01/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```




Pie charts 
- work well when the goal is to emphasize simple fractions, such as one-half, one-third, or one-quarter
- work well when we have very small datasets
- are not effective for comparing proportions

Stacked bars

- work for side-by-side comparisons of multiple conditions or in a time series


Side-by-side bar charts
- to directly compare the individual fractions to each other. 




```{r}
library(tidyverse)
library(ggplot2)
library(ggpol)
library(plyr)
```

```{r}
BRFSS <- read.csv(file="C:/Users/user/Documents/Data Science/School Programs/Data analytic, Big Datam and Predictive Analytics/Capstone Project/Submission/R codes/BRFSS2018-V4.csv", header = TRUE, sep = ",")
```

Stacked bars are used when we have only two bars in each stack

Example: to examine the difference in earning between male and female (data source: BRFSS2018)


Here is how the aggregated data look like

```{r}
df_GenderIncome <- BRFSS %>% 
  group_by(GENDER, INCOME) %>% 
  dplyr::summarise(Count =  n()) %>% 
  ddply(.(INCOME),transform, Percentage = Count*100/sum(Count)) %>% 
  mutate(pos = cumsum(Percentage) - (0.5 * Percentage))
                      
df_GenderIncome
```


```{r, fig.width= 7.5, fig.height= 7.5 }
ggplot(data = df_GenderIncome)+
  geom_bar(aes(x = INCOME, 
               y = Percentage,
               fill = GENDER),
               stat = "identity") +
   labs(y = "Relative proportion (%)", 
        x = "Salary Bracket") +
  scale_fill_manual(values = c("#A9A9A9", "#FF8C00")) +
  scale_y_continuous(breaks = seq(0, 100, 25)) +
  geom_hline(yintercept = 50,  linetype = "dashed", color = "blue")+
  theme(axis.text = element_text(size = 12, color = "black"),
        axis.title = element_text(size = 12, face = "bold"),
        legend.text = element_text(colour="black", size = 12),
        legend.position="right")
```























